Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-21 paper.tex
Syncing

\section{Introduction}

Recombination via template switching is a common feature in the evolution of
coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of
more immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.

Recombination was challenging to detect in SARS-CoV-2 early in the pandemic due
to gradual accumulation of genetic variation. In late 2020, the appearance and
spread of variants of concern (VoC) harbouring multiple characteristic
mutations increased the detectability of recombination. By searching for
genomes containing segments from different VoC, multiple recombinant lineages
have been detected, for example, between Alpha and pre-VoC
\citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and
Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for recombination
\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
transmission (more than five descendants) and monophyly
($>70\%$ bootstrap support) are eligible for a Pango lineage
designation~\citep{Rambaut2020-dw},
starting with the letter X
% FIXME: should this be a citation?
(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant Lineages'' 2021).

Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of mutations from different lineages
\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
This approach only identifies those inter-lineage recombination events
resulting in the patterns investigated. Others have inferred recombinant
genealogies to disentangle recombination from recurrent mutation, without
having to pre-specify characteristic mutations for each lineage
\citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it}
presented a phylogeny-based approach (``RIPPLES'') to identify putative
recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying motifs.
To find candidate recombinants, RIPPLES first builds a phylogeny
without assuming recombination and then scans for long branches (i.e., carrying
many mutations) and then determines if these candidates would be better
explained by recombination. Specifically, RIPPLES exhaustively breaks each
candidate sequence into segments and attempts to find more parsimonious
placements for each segment on the phylogeny. If such placements are found, the
sequence is identified as a putative recombinant. Although it enables rapid
searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2
phylogeny that accounts for only mutations, treats recombinants
\textit{post hoc}, and is an incomplete representation of the reticulate evolutionary
history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment
of recombination is possible when recombinant lineages are rare and leave few
descendants. However, this is becoming increasingly untenable with the proliferation
of recombinant lineages; for example, more than half of the sequences sampled in
February 2023 are from XBB and its descendants~\citep{Chen2022-pz}.

Inferring evolutionary trees while accounting for both mutation and
recombination is a long-standing challenge. This challenge is particularly
acute for SARS-CoV-2, given the large amount of data that has been shared, with
over 15 million sequences in the GISAID database \citep{Shu2017-hp}. It is well
known that recombination distorts phylogenies \citep{Schierup2000-fg} and
affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}.
However, standard phylogenetic methods do not account
for recombination \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}.
Alternatively, genetic ancestries of recombining lineages can be represented by
ancestral recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if},
which  capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.

Approaches to infer ARGs now exist that can scale to tens of thousands of human
genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
\cite{Wohns2022-th} reconstructed a unified whole-genome genealogy
of thousands of modern and ancient human genomes, based on over 90 million
variable sites. To achieve this, \cite{Wohns2022-th} utilised
several computational advances in ARG inference: (1) the "succinct tree
sequence" data structure that can compactly store large ARGs
\citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM)
implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to
infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient,
open-source software tools
\citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

Here, we present a preliminary version of a novel scalable methodology
(``sc2ts'', optionally pronounced ``scoots'') that builds on the advances above and
additionally takes advantage of known collection dates to infer ARGs from
millions of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2
genome sequences, we demonstrate that sc2ts can be used to simultaneously (1)
detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic
relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational
spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary
information about SARS-CoV-2 that can be efficiently accessed using "tskit"
(tree sequence toolkit) software library \citep{Kelleher2018-xc}. Sc2ts is a
promising method to automate detection of SARS-CoV-2 recombinants and to
jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at large scale.

@@

\subsection{Data preprocessing}

\subsubsection{Sequence-level preprocessing}

The findings of this study are based on sequences and metadata available on
GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
% FIXME
doi.org/10.55876/gis8.230329cd (Supplementary Table 1).
Sequences were removed
if they (1) had ambiguous collection dates;
(2) were collected before January 01, 2020 or after June 30, 2022; or
(3) were isolated from a non-human host.
Sequences were aligned to the
Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3)
using Nextclade v2.3.0~\citep{Aksamentov2021-hj}
(dataset tag 2022-07-26T12:00:00Z).
To reduce the impact of poor-quality sequences on ARG inference,
sequences were excluded if they had a ``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and
``qc.stopCodons.status'').
These excluded genomes are enriched for low-coverage, contaminated, or
misassembled sequences.

\subsubsection{Site-level preprocessing}

Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise
genome alignments were encoded as N. Problematic bases in the alignments, which
had two or more ambiguous letters or gaps within a distance of seven bases,
were masked as missing data following the protocol in
\url{https://github.com/roblanf/sarscov2phylo/}.
Masked bases were ignored when attaching a sample genome to the ARG,
but they were subsequently imputed.
Further, 481 problematic sites flagged as prone to sequencing errors or highly
homoplasic were excluded entirely
% Do we need this first URL?
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsubsection{Exclusion of ``time travellers''}

Including samples with incorrectly reported collection dates (e.g., an Alpha
sample purportedly collected in 2020 from the United States, before Alpha
appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
artefacts that distorted the topology of ARGs inferred early on in the study.
Hence, to exclude such potential ``time travellers'' we employed two filters.
The first was to remove any sequence with a collection date that pre-dated the
time to the most recent common ancestor (tMRCA) of its corresponding clade. The
tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
(downloaded on August 22, 2022) where we used the lower bound of the 95\%
confidence interval of each clade as the minimum date cut-off. The second
filter was a simple threshold on the time delay between the collection date and
submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
inferred from the MSD-filtered genome set had noticeably fewer branches with a
high number of reversions. After some preliminary analysis, a MSD threshold
value of 30 days was used when building the ARGs described herein. Future work
could estimate collection dates by adding these potential time travellers back
on the tree, allowing an automated assessment of collection date discrepancies.

\subsection{Inference}

\subsubsection{Overview of sc2ts}

\begin{figure} \centering
\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method. Sc2ts
reconstructs the genetic relationships among SARS-CoV-2 genomes by copying
samples to all possible ancestors collected at earlier time points (curved
arrows). An ARG is built by attaching new sample genomes grouped by collection
date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
Each iteration involves three stages: (1) attachment of new samples to the
growing ARG (A, B); (2) reconstruction of trees relating the samples under each
attachment node (C); and parsimony-based tree topology adjustments (D, E). In
the absence of  recombination, sc2ts infers an ARG that is a single tree
relating the samples (A). However, in the presence of recombination, sc2ts
infers an ARG that is encoded as a sequence of local trees relating segments of
the sample genomes (B), inserting a recombination node to represent a putative
recombination event. Additionally, mutation-collapsing nodes (D) and
pushed-reversion nodes (E) are inserted to make more parsimonious placements of
mutations that should be shared or should not be immediately reverted,
respectively.}
\end{figure}

We introduce an ARG inference method, summarised in
Figure~\ref{fig:overview_sc2ts}.
% Note going to rewrite this anyway, so not trying to make it flow.
It uses (1) an
efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib}
to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported
collection dates of the SARS-CoV-2 samples to determine the order of genome
copying; and (3) several parsimony-based heuristics to locally refine the
topology. In brief, this combined approach uses the likelihood model of LS to
attach sample genomes to existing nodes and to detect recombinants, but because
many of the tip samples attached daily harbour few, if any, mutational changes,
we use parsimony to resolve the relationships near the tip samples.

Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
taking advantage of the reported collection dates to incrementally grow the
ARG. The inference process occurs forward in time, attaching new genomes
aggregated by their collection date (daily batches), from the earliest to the
last collection date. Prior to attaching sample genomes, the ARG is initialised
such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
of recombination, the ARG is a single tree with no reticulation (Figure 1A);
however, in the presence of recombination, a reticulate ARG is generated, with
variation in tree topology across the genome (Figure 1B). Unlike standard
phylogenetic trees in which internal nodes represent inferred ancestors,
internal nodes in the ARGs inferred using sc2ts represent ancestors which can
be samples or inferred ancestors.

@@

\subsubsection{Brief description of the Li and Stephens model}

The LS model is a convenient probabilistic approximation to the coalescent with
recombination \citep{Hudson1983-if}, itself a generative model of a genealogical
process with recombination \citep{Li2003-ib}. Importantly, it can be implemented
as an HMM in which a focal genome is a sequence of nucleotides (observed
states) that are probabilistically emitted as an imperfect mosaic of a set of
reference genomes (hidden states). As such, an efficient HMM machinery can be
employed to make statements about parameters of interest and highlight
potential sources of error in genotyping and sequencing data. The LS model is
used in a wide variety of applications in genomics. For example, it is the
engine that drives statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
can be used to estimate mutation rates (e.g., REF), recombination rates
\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral sequence
data \citep[e.g.][]{Palmer2019-wa}.

@@

Because the HMM is memoryless, the potential parental lineages are
independently assessed at each recombinantion breakpoint. This can result in
the inference of a recombinant lineage descending immediately from three (or
even more) parental lineages. We find that often two of the three parents are
identical or similar in sequence in the segments contributed to the recombinant
lineage. A more parsimonious explanation would be to switch back to the
original parent sequence. In the future, a pragmatic solution would be to
detect such cases during the ARG inference stage and to adjust the copying
paths accordingly. Meanwhile, however, we urge caution when interpreting the
parentage of lineages with multiple breakpoints.

\subsubsection{Parameterization of the HMM implementation of Li and Stephens
model}

Recombination is inferred using an efficient HMM implementation of the LS
model. Copying paths are estimated from SARS-CoV-2 genomes sampled over time,
with younger genomes copying (i.e. inheriting segments and mutations) from
older genomes. Here, we were able to apply the LS model on millions of
SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of
the LS model that is a core component of the tsinfer software. For this
application, we have parameterised the HMM using a mismatch ratio (MMR), which
tunes how strongly mutations (i.e., no switching of template sequences) are
preferred over recombination (i.e., switching between sequences) during
copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
focal genome more likely spans at least two ancestral genomes (i.e,. the focal
genome is a putative recombinant) than a single ancestral genome. Here, the MMR
was set to three in all the HMM runs (see Discussion for future potential
development of the HMM).

\subsection{Analysis}

\subsubsection{Imputation of Pango lineage for non-sample nodes}

Non-sample nodes inserted during ARG inference had no Pango lineage assignment
from Nextclade. We imputed the Pango lineage of non-sample nodes using a
two-step procedure (Figure 3) to help assess how consistently breakpoint
intervals are estimated by the HMM runs (see below). If the Pango lineage of
the parental node was available and no mutation was immediately ancestral to
the non-sample node, then the parental node’s status was directly propagated to
the non-sample node. Otherwise, the Pango lineage of the non-sample node was
inferred by matching its full set of mutations against the Pango
lineage-defining mutations (based on 90\% consensus of the sequences analysed)
from the COVID-CG website \citep{Chen2021-zc}; \url{https://covidcg.org/};
accessed on November 04, 2022).

\subsubsection{Breakpoint intervals and HMM consistency}

When recombinant ancestry for a given genome is inferred by the LS HMM, a
breakpoint is identified as the last possible position where the copying path
must switch. Thus, the breakpoints used in the ARG represent the rightmost end
of the interval, within which the recombination event could have occurred. To
obtain the leftmost end of this interval, we rerun the HMM using mirrored
coordinates (i.e., from the 3’ end to the 5’ end).

@@

\subsubsection{Code availability}

The source code implementing the sc2ts pipeline is available via GitHub
\url{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses
various functions and algorithms implemented in tskit and tsinfer
\url{https://github.com/tskit-dev/}. The pipeline uses the SciPy
implementation of the UPGMA algorithm to build daily trees
\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
The Jupyter notebooks used to produce the results described herein are also
available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.

\section{Results}

\subsection{Inferred ARGs}

To illustrate the use of sc2ts, we employed two different sampling strategies.
First, we inferred a densely sampled but time-limited ``wide'' ARG by attaching
all the sample genomes with an MSD of 30 days, one daily batch at a time. The
inference procedure ran until the genomes sampled on June 30, 2021 were
attached to the ARG, resulting in a total of 1.27 million samples. Second, we
inferred a ``long'' ARG over a more extended timespan, by attaching at most 1,000
genomes, which were randomly subsampled from each daily batch with an MSD
within 30 days. For this ARG, the inference procedure ran until the genomes
sampled on June 30, 2022 were attached to the ARG. The long ARG covers an extra
year of samples but is more sparsely sampled. June 30, 2022 is the collection
date of the final batch of sequences which were preprocessed for the
development of sc2ts. Once sc2ts is further developed and tested, we will
attach samples collected after that date to infer a fully sampled
pandemic-scale ARG.

@@

Inferring the wide ARG required 17 days of computation time on a machine
with 128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz).
The long ARG required 23 days on a (much older) machine
with 40 threads and 256 GB RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz).
The majority of the time is spent on running the HMM to find the copying
path for each sequence, and the process is therefore highly amenable to
distributing across multiple machines.

\subsection{Backbone phylogeny}

For validation purposes, we compare the genetic relationships in the
sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain
project \citep{Hadfield2018-ef}. The NextStrain trees consist of subsamples of
SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely
restricted to non-recombinant lineages. To allow for comparison, the ARGs were
``simplified'' \citep{Kelleher2018-xc} to retain only those samples of our genealogy
that match those present in a specific NextStrain tree. These simplified
``backbone'' ARGs exhibit very little recombination (6 and 5 trees across the
genome for the wide and long ARGs, respectively).

@@

By searching for samples combining genomic segments from Alpha (B.1.1.7) and
from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
Jackson et al. (Jackson et al. 2021) reported 16 recombinant sequences among
genomes sampled in the United Kingdom up to March 7, 2021. The authors proposed
that the recombinants likely arose from eight independent recombination events
(categorised into groups A to D and four singletons).

>>>>

\section{Introduction}
Recombination via template switching is a common feature in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2 \citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing together mutations carried by different lineages, recombination plays an important role in generating genomic diversity, with recombinant lineages associated with adaptation to new host species and with the production of more immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.

Recombination was challenging to detect early in the SARS-CoV-2 pandemic due to gradual accumulation of genetic variation. In late 2020, the appearance and spread of variants of concern (VoC), designated into classes such as Alpha and Delta which harboured multiple characteristic mutations, increased the detectability of recombination. By searching for genomes containing segments from different VoC designations, multiple recombinant lineages have been detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron, with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has accelerated the rate of coinfection and the potential for recombination \citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward transmission (more than five descendants) and monophyly
($>70\%$ bootstrap support) are eligible for a Pango lineage designation~\citep{Rambaut2020-dw}, starting with the letter X
% FIXME: should this be a citation?
(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant Lineages'' 2021).

Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that combine specific subsets of mutations from different lineages \citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}. This approach only identifies those inter-lineage recombination events resulting in the patterns investigated. Others have inferred recombinant genealogies to disentangle recombination from recurrent mutation, without having to pre-specify characteristic mutations for each lineage \citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it} presented a phylogeny-based approach (``RIPPLES'') to identify putative recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find candidate recombinants, RIPPLES first builds a phylogeny without assuming recombination and then scans for long branches (i.e., carrying many mutations) and then determines if these candidates would be better explained by recombination. Specifically, RIPPLES exhaustively breaks each candidate sequence into segments and attempts to find more parsimonious placements for each segment on the phylogeny. If such placements are found, the sequence is identified as a putative recombinant. Although it enables rapid searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that accounts for only mutations, treats recombinants \textit{post hoc}, and is an incomplete representation of the reticulate evolutionary history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment of recombination is possible when recombinant lineages are rare and leave few descendants. However, this is becoming increasingly untenable with the proliferation of recombinant lineages; for example, more than half of the sequences sampled in February 2023 are from XBB and its descendants~\citep{Chen2022-pz}. This also means that future evolution of SARS-CoV-2 is likely to involve multiple sequential recombination events on top of existing recombinant lineages, creating a highly reticulated genealogy.

Inferring evolutionary trees while accounting for both mutation and recombination is a long-standing computational challenge. This challenge is particularly acute for SARS-CoV-2, given the large amount of data that has been shared, with over 15 million sequences in the GISAID database \citep{Shu2017-hp}. It is well known that recombination distorts phylogenies \citep{Schierup2000-fg} and affects the results of downstream analysis, e.g., inference of selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do not account for recombination \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Alternatively, genetic ancestries of recombining lineages can be represented by ancestral recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if}, which capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.

Approaches to infer ARGs now exist that can scale to tens of thousands of human genomes and beyond  \citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which deal with levels of sequential recombination far in excess of those seen in viral phylogenies. For example, using over 90 million variable sites \citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands of modern and ancient human genomes, accounting for millions of recombination events (over 5 million "local trees"). To achieve this, \cite{Wohns2022-th} took advantage of several computational advances in ARG inference: (1) the "succinct tree sequence" data structure that can compactly store large ARGs \citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM) implementation of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient, open-source software tools \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

Here, we present a preliminary version of a novel scalable methodology
(``sc2ts'', optionally pronounced ``scoots'') that builds on the advances above and additionally takes advantage of known collection dates to infer ARGs from millions of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences, we demonstrate that sc2ts can be used to simultaneously (1) detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary information about SARS-CoV-2 that can be efficiently accessed using "tskit" (tree sequence toolkit) software library \citep{Kelleher2018-xc}. Sc2ts is a promising method to automate detection of SARS-CoV-2 recombinants and to jointly model recombinational and mutational processes shaping the genomic diversity of SARS-CoV-2 at scale.

@@

\subsection{Data preprocessing}
\subsubsection{Sequence-level preprocessing}
The findings of this study are based on sequences and metadata available on GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
% FIXME
doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed if they (1) had ambiguous collection dates; (2) were collected before January 01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host. Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag 2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG inference, sequences were excluded if they had a ``bad'' quality control status in any of the four Nextclade columns (``qc.missingData.status'', ``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status''). These excluded genomes are enriched for low-coverage, contaminated, or misassembled sequences.

\subsubsection{Site-level preprocessing}
Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N. Problematic bases in the alignments, which had two or more ambiguous letters or gaps within a distance of seven bases, were masked as missing data following the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases were ignored when attaching a sample genome to the ARG, but they were subsequently imputed. Further, 481 problematic sites flagged as prone to sequencing errors or highly homoplasic were excluded entirely
% Do we need this first URL?
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473}; \url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsubsection{Exclusion of ``time travellers''}
Including samples with incorrectly reported collection dates (e.g., an Alpha sample purportedly collected in 2020 from the United States, before Alpha appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce artefacts that distorted the topology of ARGs inferred early on in the study. Hence, to exclude such potential ``time travellers'' we employed two filters. The first was to remove any sequence with a collection date that pre-dated the time to the most recent common ancestor (tMRCA) of its corresponding clade. The tMRCA for each clade was obtained from a Nextstrain GISAID reference tree (downloaded on August 22, 2022) where we used the lower bound of the 95\% confidence interval of each clade as the minimum date cut-off. The second filter was a simple threshold on the time delay between the collection date and submission date (maximum submission delay, or MSD). Indeed, we found that ARGs inferred from the MSD-filtered genome set had noticeably fewer branches with a high number of reversions. After some preliminary analysis, a MSD threshold value of 30 days was used when building the ARGs described herein. Future work could estimate collection dates by adding these potential time travellers back on the tree, allowing an automated assessment of collection date discrepancies.

\subsection{Inference}
\subsubsection{Overview of sc2ts}
\begin{figure} \centering
\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method. Sc2ts reconstructs the genetic relationships among SARS-CoV-2 genomes by copying samples to all possible ancestors collected at earlier time points (curved arrows). An ARG is built by attaching new sample genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day. Each iteration involves three stages: (1) attachment of new samples to the growing ARG (A, B); (2) reconstruction of trees relating the samples under each attachment node (C); and parsimony-based tree topology adjustments (D, E). In the absence of  recombination, sc2ts infers an ARG that is a single tree relating the samples (A). However, in the presence of recombination, sc2ts infers an ARG that is encoded as a sequence of local trees relating segments of the sample genomes (B), inserting a recombination node to represent a putative recombination event. Additionally, mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to make more parsimonious placements of mutations that should be shared or should not be immediately reverted, respectively.}
\end{figure}

We introduce an ARG inference method, summarised in
Figure~\ref{fig:overview_sc2ts}.
% Note going to rewrite this anyway, so not trying to make it flow.
It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported collection dates of the SARS-CoV-2 samples to determine the order of genome copying; and (3) several parsimony-based heuristics to locally refine the topology. In brief, this combined approach uses the likelihood model of LS to attach sample genomes to existing nodes and to detect recombinants, but because many of the daily tip samples harbour few, if any, mutational changes, we use parsimony to resolve their close relationships.

Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by taking advantage of the reported collection dates to incrementally grow the ARG. The inference process occurs forward in time, attaching new genomes aggregated by their collection date (daily batches), from the earliest to the last collection date. Prior to attaching sample genomes, the ARG is initialised such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence of recombination, the ARG is a single tree with no reticulation (Figure 1A); however, in the presence of recombination, a reticulate ARG is generated, with variation in tree topology across the genome (Figure 1B). Unlike standard phylogenetic trees in which internal nodes always represent inferred ancestors, internal nodes in the sc2ts ARGs represent sequences which can either be inferred ancestors or known samples.

@@

\subsubsection{Brief description of the Li and Stephens model} 
The LS model is a convenient probabilistic approximation to the coalescent with recombination \citep{Hudson1983-if}, itself a generative model of a genealogical process with recombination \citep{Li2003-ib}. Importantly, it can be implemented as an HMM in which a focal genome is a sequence of nucleotides (observed states) that are probabilistically emitted as an imperfect mosaic of a set of reference genomes (hidden states). As such, an efficient HMM machinery can be employed to make statements about parameters of interest and highlight potential sources of error in genotyping and sequencing data. The LS model is used in a wide variety of applications in genomics. For example, it is the engine that drives statistical genotype phasing and imputation \citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it can be used to estimate mutation rates (e.g., REF), recombination rates \citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral sequence data \citep[e.g.][]{Palmer2019-wa}.

@@

Because the HMM is memoryless, the potential parental lineages are independently assessed at each recombination breakpoint. This can result in the inference of a recombinant lineage descending immediately from three (or even more) parental lineages. We find that often two of the three parents are identical or similar in sequence in the segments contributed to the recombinant lineage. A more parsimonious explanation would be to switch back to the original parent sequence. In the future, a pragmatic solution would be to detect such cases during the ARG inference stage and to adjust the copying paths accordingly (see ``Improvements'', below). Meanwhile, however, we urge caution when interpreting the parentage of lineages with multiple breakpoints.

\subsubsection{Parameterization of the HMM implementation of Li and Stephens model} 
Recombination is inferred using an efficient HMM implementation of the LS model. Copying paths are estimated from SARS-CoV-2 genomes sampled over time, with younger genomes copying (i.e. inheriting segments and mutations) from older genomes. Here, we were able to apply the LS model on millions of SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of the LS model that is a core component of the tsinfer software. For this application, we have parameterised the HMM using a mismatch ratio (MMR), which tunes how strongly mutations (i.e., no switching of template sequences) are preferred over recombination (i.e., switching between sequences) during copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a focal genome more likely spans at least two ancestral genomes (i.e,. the focal genome is a putative recombinant) than a single ancestral genome. Here, the MMR was set to three in all the HMM runs (see ``Improvements'' for future potential development of the HMM).

\subsection{Analysis} 
\subsubsection{Imputation of Pango lineage for non-sample nodes} 
Non-sample nodes inserted during ARG inference had no Pango lineage assignment from Nextclade. We imputed the Pango lineage of non-sample nodes using a two-step procedure (Figure 3) to help assess how consistently breakpoint intervals are estimated by the HMM runs (see below). If the Pango lineage of the parental node was available and no mutation was immediately ancestral to the non-sample node, then the parental node’s status was directly propagated to the non-sample node. Otherwise, the Pango lineage of the non-sample node was inferred by matching its full set of mutations against the Pango lineage-defining mutations (based on 90\% consensus of the sequences analysed) from the COVID-CG website \citep{Chen2021-zc}; \url{https://covidcg.org/}; accessed on November 04, 2022).

\subsubsection{Breakpoint intervals and HMM consistency}
 When recombinant ancestry for a given genome is inferred by the LS HMM, a breakpoint is identified as the last possible position where the copying path must switch. Thus, the breakpoints used in the ARG represent the rightmost end of the interval, within which the recombination event could have occurred. To obtain the leftmost end of this interval, we rerun the HMM using mirrored coordinates (i.e., from the 3’ end to the 5’ end).

@@

\subsubsection{Code availability}
 The source code implementing the sc2ts pipeline is available via GitHub \url{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses various functions and algorithms implemented in tskit and tsinfer \url{https://github.com/tskit-dev/}. The pipeline uses the SciPy implementation of the UPGMA algorithm to build daily trees \url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}. The Jupyter notebooks used to produce the results described herein are also available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.

\section{Results}
\subsection{Inferred ARGs} 
To illustrate the use of sc2ts, we employed two different sampling strategies. First, we inferred a densely sampled but time-limited ``wide'' ARG by attaching all the sample genomes with an MSD of 30 days, one daily batch at a time. The inference procedure ran until the genomes sampled on June 30, 2021 were attached to the ARG, resulting in a total of 1.27 million samples. Second, we inferred a ``long'' ARG over a more extended timespan, by attaching at most 1,000 genomes, which were randomly subsampled from each daily batch with an MSD within 30 days. For this ARG, the inference procedure ran until the genomes sampled on June 30, 2022 were attached to the ARG. The long ARG covers an extra year of samples but is more sparsely sampled. June 30, 2022 is the collection date of the final batch of sequences which were preprocessed for the development of sc2ts. Once sc2ts is further developed and tested, we will attach samples collected after that date to infer a fully sampled pandemic-scale ARG.

@@

Inferring the wide ARG required 17 days of computation time on a machine with 128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). Inferring the long ARG required 23 days on a (much older) machine with 40 threads and 256 GB RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz). The majority of the time is spent on running the HMM to find the copying path for each sequence, and the process is therefore highly amenable to distributing across multiple machines.

\subsection{Backbone phylogeny} 
For validation purposes, we compare the genetic relationships in the sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain project \citep{Hadfield2018-ef}. The NextStrain trees consist of subsamples of SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely restricted to non-recombinant lineages. To allow for comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to retain only those samples of our genealogy that match those present in a specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very little recombination (6 and 5 trees across the genome for the wide and long ARGs, respectively).

@@

By searching for samples combining genomic segments from Alpha (B.1.1.7) and from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations, \citet{Jackson2021-ik} reported 16 recombinant sequences among genomes sampled in the United Kingdom up to March 7, 2021. The authors proposed that the recombinants likely arose from eight independent recombination events (categorised into groups A to D and four singletons).
